---
title: "MNIST example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MNIST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Prepare the MNIST dataset

```{r setup, warning = F}
library(dslabs)
set.seed(1)

# load dataset
mnist <- read_mnist()

# sample sizes
n_labeled <- 500
n_unlabeled <- 5000

# known vs. unknown classes
known_classes <- 0:4
all_classes <- sort(unique(mnist$train$labels))
labeled_train <- as.vector(sapply(known_classes, 
                        function(x){
                          return(sample(which(mnist$train$labels == x),
                                        n_labeled / length(known_classes)))
                        }))
unlabeled_train <- as.vector(sapply(all_classes, 
                        function(x){
                          return(sample(setdiff(which(mnist$train$labels == x), labeled_train),
                                        n_unlabeled / length(all_classes)))
                        }))

X <- mnist$train$images[union(labeled_train, unlabeled_train),]
y <- c(mnist$train$labels[labeled_train], rep(NA, length(unlabeled_train)))
ytrue <- mnist$train$labels[union(labeled_train, unlabeled_train)]

summary(as.factor(y))
summary(as.factor(ytrue))
```

```{r real-world data}
# input dataset for the model
data <- as.data.frame(cbind(X, y))

# model formula
formula <- y ~ . - 1
```

## Train SSC-UC model

The SSC-UC model is called via <tt>SSCUC</tt> and returns a <tt>BayesClassifier</tt> object with known and unknown classes. In this case, a Bayes classifier is used, specified by the argument <tt>naive=TRUE</tt>.

```{r train}
library(SSCUC)

# train model
model <- SSC(formula, data, naive = TRUE,
             prior = "uniform")
summary(model)
```

## Predict and evaluate unlabeled data using SSC-UC model

```{r predict}
# predict on test dataset
test <- as.data.frame(cbind(mnist$test$images, 
                            y = mnist$test$labels))

pred <- predict(model, newdata = test, type = "class")
```

### Evaluate in full confusion matrix (multiple unknown classes)

```{r evaluate1}
library(caret)
library(knitr)
library(ggplot2)

# specify levels
levels <- sort(union(unique(pred), unique(test$y)))

# compute confusion matrix
cm <- confusionMatrix(
    reference = factor(test$y, levels = levels),
    data = factor(pred, levels = levels))

# print macro-averaged F1 score
cat("Macro-averaged F1 score: ", mean(cm$byClass[,"F1"], na.rm = TRUE))
```

```{r evaluate2, fig.height = 6, fig.width = 6, fig.align = 'center', echo=FALSE}
df <- data.frame(
  value = as.vector(cm$table[paste0(sort(unique(pred))),paste0(sort(unique(test$y)))]),
  predicted = rep(sort(unique(pred)), length(unique(test$y))),
  true = rep(sort(unique(test$y)), each = length(unique(pred)))
)
ggplot(df, aes(x = true, y = predicted, fill = value)) +
  geom_tile(alpha = 0.8) +
  geom_text(aes(label = value)) +
  theme_bw() + 
  theme(legend.position = "bottom") + 
  labs(x = "true label",
       y = "predicted label",
       title = "SSC-UC results on MNIST dataset",
       subtitle = "10-class dataset with 5 unknown classes - confusion matrix",
       fill = "absolute count") + 
  scale_fill_gradient(low = "white", high = "black", limits = c(0, max(df$value))) + 
  scale_x_continuous(breaks = 0:9)
```

### Visualization of results
```{r plot1, fig.height = 5, fig.width = 10, fig.align = 'center', echo = F}
library(ggplot2)
library(latex2exp)

mean <- sd <- matrix(NA, ncol = length(model), nrow = length(model$all.features), 
         dimnames = list(model$all.features, levels(model)))

mean[dimnames(model),] <- sapply(levels(model),function(x){return(model$param[[x]]$mu)})
sd[dimnames(model),] <- sapply(levels(model), function(x){return(sqrt(diag(model$param[[x]]$Sigma)))})

df <- data.frame(
  x1 = rep(1:28, 28 * length(model)),
  x2 = rep(1:28, each = 28),
  class = rep(levels(model), each = length(model$all.features)),
  type = rep(c("mu", "sigma"), each = 784 * length(model)),
  value = c(mean, as.vector(sd))
)

appender <- function(string) ifelse(string %in% c("mu", "sigma"), TeX(paste0("$\\",string,"$")), string)
ggplot(df, aes(x = x1, y = -x2)) + 
  geom_tile(aes(fill = value)) +
  facet_grid(type~class, 
             labeller = as_labeller(appender, default = label_parsed)) + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL) +
  labs(fill = "", 
       x = "",
       y = "",
       title = "SSC-UC results on MNIST dataset",
       subtitle = TeX("10-class dataset with 5 unknown classes (class mean $\\mu$ and variance $\\sigma$)"))
```
