---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Prepare a sample dataset for Naive Bayes

In the first step, we simulate a dataset with the following specifications:

- a total number of <tt>5</tt> classes, 2 of which are known and 3 are unknown, respectively
- <tt>n_labeled = 300</tt> labeled samples, evenly distributed over the 2 known classes
- <tt>n_unlabeled = 3000</tt> unlabeled samples, evenly distributed over the 5 classes
- a dimension of <tt>n_feats = 2</tt> features (to facilitate representation)

Each class $i$ is represented by a bivariate Gaussian distribution with mean vector $\mu = (\mu_1^{(i)}, \mu_2^{(i)})$ and covariance matrix $$\Sigma = \left(\begin{array}{cc}1 & 0 \\ 0 & 1\end{array}\right)$$ (identical for all classes). The known classes are 0 and 1.

```{r setup}
library(mvtnorm)
set.seed(1)

# sample sizes
n_labeled <- 300
n_unlabeled <- 3000
n_feats <- 2

# mean values and covariance matrices
mu <- rbind(
  c(-5, 10),
  c(0, 5),
  c(-5, 0),
  c(-10,5),
  c(10,-10)
)
rownames(mu) <- 0:4
sigma <- diag(rep(1,n_feats))

print(mu)
print(sigma)
```

Given the data specifications, we simulate from the bivariate Gaussian distribution to generate the dataset given the class vector. We discriminate between the true class vector <tt>y_true</tt> containing the labels <tt>0-4</tt> of all samples, and the model input class vector <tt>y</tt> containing a true label only for the labeled data, and <tt>NA</tt>, otherwise.

```{r simulate data}
# specify number of labeles / unlabeled samples per class
labeled_classes <- rep(0:1, each = n_labeled / 2)
unlabeled_classes <- rep(0:4, each = n_unlabeled / 5)
num_sample <- c(table(labeled_classes), table(unlabeled_classes))

# simulate X, ytrue and y
X <- c()
for(i in 1:length(num_sample)){
  X <- rbind(X, rmvnorm(num_sample[i], mu[names(num_sample)[i],], sigma))
}
X <- cbind(X, 1)
y <- rep(c(1:2, NA), c(table(labeled_classes), length(unlabeled_classes)))
ytrue <- rep(names(num_sample), num_sample)
colnames(X) <- paste0("x", 1:3)

# summaries of the model input class vector y, and the true class vector ytrue
summary(as.factor(y))
summary(as.factor(ytrue))
```

Using the model input class vector <tt>y</tt> and the simulated feature matrix <tt>X</tt>, we generate the input dataset and the formula for the model.

```{r simulate data2}
# input dataset for the model
data <- as.data.frame(cbind(X, y))

# model formula
formula <- y ~ x1 + x2 + x3 - 1

# simulated data
head(data)
```

The simulated dataset is given by the following scatterplot:

```{r plot0, fig.height = 6, fig.width = 6, echo = FALSE, fig.align = 'center'}
library(ggplot2)
ggplot(rbind(subset(data, is.na(y)), subset(data, !is.na(y))),
       aes(x = x1, y = x2, color = as.factor(y))) + 
  geom_point(alpha = 0.5) + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  labs(color = "model input class label", 
       title = "simulated dataset for SSC-UC",
       subtitle = "5-class dataset with 3 unknown classes")
```

## Train SSC-UC model

The SSC-UC model is called via <tt>SSCUC</tt> and returns a <tt>BayesClassifier</tt> object with known and unknown classes. In this case, a naive Bayes classifier is used, specified by the argument <tt>naive=TRUE</tt>.

```{r train}
library(SSCUC)

# train model
model <- SSC(formula, data, naive = TRUE)
summary(model)
```

## Predict and evaluate unlabeled data using SSC-UC model

Class labels for the unlabeled data in the dataset are obtained using the <tt>predict</tt> function with argument <tt>type="class"</tt>.

```{r predict}
# predict on unlabeled data
pred <- predict(model, newdata = subset(data, is.na(y)), type = "class")
```

### Evaluate in full confusion matrix (multiple unknown classes)

The full confusion matrix contains all the classes modeled by the <tt>BayesClassifier</tt>:

- the class labels <tt>0-4</tt> from the data specification as reference labels (not that, however, only classes 0 and 1 are known to the model a-priori),
- the class labels for known classes (<tt>0-1</tt>) and unknown classes (<tt>U1-U4</tt>) as predicted labels.

```{r evaluate1}
library(caret)
library(knitr)

# specify levels
levels <- sort(union(unique(pred), unique(ytrue)))

kable(confusionMatrix(
  reference = factor(ytrue[is.na(y)], levels = levels), 
  data = factor(pred, levels = levels))$table[sort(unique(pred)),sort(unique(ytrue))]
)
```

The following plot shows the unlabeled data with their predicted class labels:

```{r plot1, fig.height = 6, fig.width = 6, fig.align = 'center'}
library(ggplot2)
ggplot(cbind(subset(data, is.na(y)), pred = pred), 
       aes(x = x1, y = x2, color = as.factor(pred))) + 
  geom_point(alpha = 0.5) + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  labs(color = "prediction", 
       title = "simulated dataset for SSC-UC",
       subtitle = "prediction of all known and unknown classes")
```

### Evaluate in reduced confusion matrix (all unknown classes as class "U")

As a second evaluation step, we can summarize all "unknown" classes detected by SSC-UC into one class "U" (unknown). Thereby, we reduce the number of labels in the confusion matrix to <tt>0-1</tt> and <tt>U</tt>.

```{r evaluate2}
# set all new class labels to "U"
pred[!pred %in% unique(ytrue)] <- "U"

# specify levels
levels <- sort(union(unique(pred), unique(ytrue)))

kable(confusionMatrix(
  reference = factor(ytrue[is.na(y)], levels = levels), 
  data = factor(pred, levels = levels))$table[sort(unique(pred)),sort(unique(ytrue))]
)
```

The following plot shows the unlabeled data with their predicted class labels, when considering all unknown classes as one class.

```{r plot2, fig.height = 6, fig.width = 6, fig.align = 'center'}
library(ggplot2)
ggplot(cbind(subset(data, is.na(y)), pred = pred), 
       aes(x = x1, y = x2, color = as.factor(pred))) + 
  geom_point(alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "bottom") + 
  labs(color = "prediction", 
       title = "simulated dataset for SSC-UC",
       subtitle = "prediction of known versus unknown classes")
```
